Computing the {\em least-commitment} action model from state observations

### 2018-11-30 doc/ijcai-19-conformant/conformant-learning.tex
adding comments by Eva and Diego from todays meeting

A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,s_0)$, meaning that $\forall i, s_i^o \subseteq s_i$. In practice, the number of observed states, $m$, ranges from 1 (the initial state, at least), to $|\pi|+1$, and the observed intermediate states will comprise a number of fluents between $[1,|F|]$.

In other words, we assume there is a bijective monotone mapping between trajectories and observations~\cite{ramirez2009plan}, thus also granting the inverse consistency relationship (the trajectory is a superset of the observation). Therefore, transiting between two consecutive observed states in $\mathcal{O}(\tau)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite. In other words, having $\mathcal{O}(\tau)$ does not imply knowing the actual length of $\pi$.

\begin{definition}[Explaning a $\mathcal{O}(\tau)$ observation]
Given a {\em classical planning problem} $P$ and a sequence of partially observed states $\mathcal{O}(\tau)$, we say that $P$ {\em explains the observation} (denoted $P\mapsto\mathcal{O}(\tau)$) iff there exists a solution plan $\pi$ for $P$ that is consistent with $\mathcal{O}(\tau)$. If $\pi$ is also optimal, we say that $\pi$ is the {\em best explanation} for $\mathcal{O}(\tau)$. 
\end{definition}

@@

\subsection{The {\em least-commitment} action model}
The task of computing the {\em least-commitment} action model from a sequence of state observations is defined as $\tup{P,\mathcal{O}(\tau)}$:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a classical planning problem where $A[\cdot]$ is a set of actions s.t. the semantics of each $a\in A[\cdot]$ is unknown; i.e. the corresponding $\tup{\rho,\theta}$ functions are undefined. The set of goals $G$ can also be unknown.
\item $\mathcal{O}(\tau)$ is a sequence of partial states that results from the observation of a trajectory $\tau(\pi,s_0)$ produced by the execution of certain unknown plan $\pi$ that solves $P$.
\end{itemize}

Before formalizing the solution to this task, i.e. the {\em least-commitment} action model, we introduce several necessary definitions. We first start defining a {\em partially specified action model} inspired by the notion of {\em incomplete (annotated) model}~\cite{sreedharan2018handling}.
\begin{definition}[Partially specified action model]
Given a set of actions $A[\cdot]$ and a set of fluents $F$ then, a {\em partially specified action model} $M$ is a set of possible models for the actions in $A[\cdot]$ such that: (1), any model $\mathcal{M}\in M$ is a definition of the $\tup{\rho,\theta}$ functions of every action in $A[\cdot]$ and (2), for every $\mathcal{M}\in M$ the $\tup{\rho,\theta}$ functions are defined in the set of state variables $F$. (Note that if $M$ is a singleton it represents a {\em fully specified action model}).
\end{definition}

Now we are ready to define the {\em least-commitment} action model for an observation $\mathcal{O}(\tau)$.
\begin{definition}[The {\em least-commitment} action model]
Given a $\tup{P,\mathcal{O}(\tau)}$ task and the {\em partially specified action model} $M$ that represents the full space of possible action models for the actions in $A[\cdot]\in P$, then the {\em least-commitment} action model is another {\em partially specified action model} that represents the largest subset of models $M^*\subseteq M$ such that every model $\mathcal{M}\in M^*$ {\em explains} the input observation.
\end{definition}

@@

{\em A \strips\ action schema} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema. Let be $\Psi$ the set of {\em predicates} that shape the propositional state variables $F$, and a list of {\em parameters} $pars(\xi)$. The set of elements that can appear in $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the \strips\ action schema $\xi$ is given by FOL interpretations of $\Psi$ over the parameters $pars(\xi)$. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For instance, in the {\em blocksworld} the ${\mathcal I}_{\Psi,\xi}$ set contain eleven elements for the {\tt stack} schemata, ${\mathcal I}_{\Psi,stack}$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$), on($v_1,v_2$), on($v_2,v_1$), on($v_2,v_2$)\}}.

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, the space of possible \strips\ schemata is constrained by constraints of three kinds:
\begin{enumerate}
\item {\em Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. 
\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this particular domain, a block cannot be on top of itself. As a rule of thumb, {\it state invariants} constraining the possible states of a given planning domain belong to this second class of constraints~\cite{fox1998automatic}. 
\item {\em Observation constraints}. A sequence of state observations $\mathcal{O}(\tau)$ is \emph{semantic knowledge} that constraints further the space of possible action schemata. 
\end{enumerate}

\subsection{Computing the {\em least-commitment} model via conformant planning}
Given the set of actions $A[\cdot]$ (with unknown $\tup{\rho,\theta}$ functions) and a sequence of partial states $\mathcal{O}(\tau)=\tup{s_0^o,s_1^o \ldots , s_m^o}$, we can build the classical planning problem $P_\mathcal{O}=\tup{F,A[\cdot],s_0^o,s_m^o}$. In other words, a classical planning problem such that $\mathcal{O}(\tau)$ is the observation of a $\tau(\pi,I)$ trajectory that solves that problem.

In this section we show that starting from a $\tup{P_\mathcal{O},\mathcal{O}(\tau)}$ task we can build a {\em conformant planning problem} $P_c$ whose solution induces the {\em least-commitment} action model for the input observation $\mathcal{O}(\tau)$. In more detail, we build a {\em conformant planning problem} $P_c=\tup{F_c,A_c,\Upsilon,G}$ such that:
\begin{itemize}
\item The set of fluents $F_c$ extends $F$ with two new sets of fluents:
      \begin{itemize}
      \item $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\mathcal{O}(\tau)$ where the action model is validated
      \item Fluents {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} (where $e\in{\mathcal I}_{\Psi,\xi}$) implementing a propositional encoding of the {\em preconditions}, {\em negative}, and {\em positive} effects of an action schema $\xi$. Our encoding exploits the syntactic constraint of \strips\ so, if {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a negative effect in $\xi$ while if $pre\_e\_\xi$ does not hold but {\tt\small eff\_e\_$\xi$} holds, it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a positive effect in $\xi$. Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema.
      \end {itemize}

@@

      \item {\em Editable} actions whose semantics is given by the value of {\tt\small pre\_e\_$\xi$}, {\tt\small eff\_e\_$\xi$} fluents at the current state. Figure~\ref{fig:editable} shows the PDDL encoding of an {\em editable} {\tt\small stack(?v1,?v2)} schema. Note that this editable schema when {\tiny\tt (pre\_holding\_v1\_stack) (pre\_clear\_v2\_stack) (eff\_holding\_v1\_stack) (eff\_clear\_v2\_stack) (eff\_clear\_v1\_stack) (eff\_handempty\_stack) (eff\_on\_v1\_v2\_stack)} hold, then it behaves exactly as the original PDDL schema defined in Figure~\ref{fig:propositional}. Formally, given an operator schema $\xi\in\mathcal{M}$ its {\em editable} version is:
\begin{small}  
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi}})=&\{pre\_e\_\xi\implies e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}\\
\cond(\mathsf{editable_{\xi}})=&\{pre\_e\_\xi, eff\_e\_\xi\}\rhd\{\neg e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}},\\
&\{\neg pre\_e\_\xi, eff\_e\_\xi\}\rhd\{e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}.
\end{align*}
\end{small}

@@

\item The clauses in $\Upsilon$ comprises:
      \begin{enumerate}
      \item The {\em unit clauses} given by the fluents that hold in the initial state $I=s_0$ and $mode_{commit}$ set to true.
      \item The clauses representing that the actual value of fluents {\tt\small pre\_e\_$\xi$}, {\tt\small eff\_e\_$\xi$} is unknown. In other words, that any model from the \strips\ space of models (following the previously mentioned syntactic constraints) can initially be part of the {\em least-commitment} action model. Formally, for every $\xi$ and $e\in{\mathcal I}_{\Psi,\xi}$, then $\Upsilon$ includes these two clauses:
            \begin{itemize}
            \item {\tt\small pre\_e\_$\xi$} xor {\tt\small $\neg$pre\_e\_$\xi$}.
            \item {\tt\small eff\_e\_$\xi$} xor {\tt\small $\neg$eff\_e\_$\xi$}.           
            \end{itemize}
      \end{enumerate}
One can also add here clauses that encode {\em domain-specific constraints} (as mentioned in the previous section) to make the conformant planning problem easier to be solved for a particular domain. 
\item The new goals are $G_c=\{test_m\}$.
\end{itemize}

@@

@@

\section{Conclusions}
~\cite{SternJ17}

%% The file named.bst is a bibliography style file for BibTeX 0.99c
\bibliographystyle{named}
\bibliography{planlearnbibliography.bib}

>>>>

A partially observable state $s_i^o$, {\small $0<i<m$}, is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,s_0)$, meaning that $\forall i, s_i^o \subseteq s_i$. In practice, the number of observed states $m$, ranges from 1 (the initial state, at least), to $|\pi|+1$, and the observed intermediate states will comprise a number of fluents between $[1,|F|]$.

In other words, we assume there is a {\em bijective monotone mapping} between trajectories and observations~\cite{ramirez2009plan}, thus also granting the inverse consistency relationship (the trajectory is a superset of the observation). Therefore, transiting between two consecutive observed states in $\mathcal{O}(\tau)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite. In other words, having $\mathcal{O}(\tau)$ does not imply knowing the actual length of $\pi$.

\begin{definition}[Explaning a $\mathcal{O}(\tau)$ observation]
Given a {\em classical planning problem} $P$ and a sequence of partially observed states $\mathcal{O}(\tau)$, we say that a plan $\pi$ {\em explains the observation} (denoted $\pi\mapsto\mathcal{O}(\tau)$) iff $\pi$ is a solution for $P$ that is consistent with $\mathcal{O}(\tau)$. If $\pi$ is also optimal, we say that $\pi$ is the {\em best explanation} for $\mathcal{O}(\tau)$. 
\end{definition}

@@

\subsection{The {\em least-commitment} action model}
The task of computing the {\em least-commitment} action model from a sequence of state observations is defined as $\tup{\Phi,\mathcal{O}(\tau)}$:
\begin{itemize}
\item $\Phi=\tup{F,A[\cdot]}$ is a {\em classical planning frame} where $A[\cdot]$ is a set of actions s.t. the semantics of each $a\in A[\cdot]$ is unknown; i.e. the corresponding $\tup{\rho,\theta}$ functions are undefined. 
\item $\mathcal{O}(\tau)$ is a sequence of partial states that results from the partial observation of a trajectory $\tau(\pi,s_0)$ that is defined within the {\em classical planning frame} $\Phi$.
\end{itemize}

Before formalizing the solution to this task, i.e. the {\em least-commitment} action model, we introduce several necessary definitions. Given a {\em classical planning frame} $\Phi=\tup{F,A[\cdot]}$ and a sequence of partial states $\mathcal{O}(\tau)=\tup{s_0^o,s_1^o \ldots , s_m^o}$, we can build the classical planning problem $P_\mathcal{O}=\tup{F,A[\cdot],s_0^o,s_m^o}$. We say that an action model $\mathcal{M}$ is a definition of the $\tup{\rho,\theta}$ functions of every action in $A[\cdot]$. Further, we say that a model $\mathcal{M}$ {\em explains} a sequence of observations $\mathcal{O}(\tau)$ iff there exists a solution plan for $P_\mathcal{O}$ when the $\tup{\rho,\theta}$ functions of the actions in $P_\mathcal{O}$ are given by $\mathcal{M}$.

Now, we define a {\em partially specified action model} inspired by the notion of {\em incomplete (annotated) model}~\cite{sreedharan2018handling}.
\begin{definition}[Partially specified action model]
Given a {\em classical planning frame} $\Phi=\tup{F,A[\cdot]}$ then, a {\em partially specified action model} $M$ is a set of possible models for the actions in $A[\cdot]$ such that: (1), any model $\mathcal{M}\in M$ is a definition of the $\tup{\rho,\theta}$ functions of every action in $A[\cdot]$ and (2), for every $\mathcal{M}\in M$ the $\tup{\rho,\theta}$ functions are defined in the set of state variables $F$. (Note that if $M$ is a singleton it represents a {\em fully specified action model}).
\end{definition}

Now we are ready to define the {\em least-commitment} action model for an observation $\mathcal{O}(\tau)$.
\begin{definition}[The {\em least-commitment} action model]
Given a $\tup{\Phi,\mathcal{O}(\tau)}$ task and the {\em partially specified action model} $M$ that represents the full space of possible action models for the actions in $A[\cdot]\in \Phi$, then the {\em least-commitment} action model is another {\em partially specified action model} that represents the largest subset of models $M^*\subseteq M$ such that every model $\mathcal{M}\in M^*$ {\em explains} the input observation.
\end{definition}

@@

{\em A \strips\ action schema} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema. Let be $\Psi$ the set of {\em predicates} that shape the propositional state variables $F$, and a list of {\em parameters} $pars(\xi)$. The set of elements that can appear in $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the \strips\ action schema $\xi$ is given by FOL interpretations of $\Psi$ over the parameters $pars(\xi)$. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For instance, in the {\em blocksworld} the ${\mathcal I}_{\Psi,\xi}$ set contain eleven elements for the {\small \tt stack($v_1$,$v_2$)} schemata, ${\mathcal I}_{\Psi,stack}$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$), on($v_1,v_2$), on($v_2,v_1$), on($v_2,v_2$)\}}.

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, the space of possible \strips\ schemata is bounded by constraints of three kinds:
\begin{enumerate}
\item {\em Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. 
\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. As a rule of thumb, {\it state invariants} constraining the possible states of a given planning domain belong to this second class of constraints~\cite{fox1998automatic}. 
\item {\em Observation constraints}. A sequence of state observations $\mathcal{O}(\tau)$ depicts {\em semantic knowledge} that constraints further the space of possible action schemata. 
\end{enumerate}

\subsection{Computing the {\em least-commitment} model via conformant planning}
Given a task $\tup{\Phi,\mathcal{O}(\tau)}$, this section shows that we can build a {\em conformant planning problem} $P_c$ whose solution induces the {\em least-commitment} action model for the input observation $\mathcal{O}(\tau)$. In more detail, we build a {\em conformant planning problem} $P_c=\tup{F_c,A_c,\Upsilon,G_c}$ such that:
\begin{itemize}
\item The set of fluents $F_c$ extends $F$ with two new sets of fluents:
      \begin{itemize}
      \item $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\mathcal{O}(\tau)$ where the action model is validated
      \item Fluents {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} (where $e\in{\mathcal I}_{\Psi,\xi}$) implementing a propositional encoding of the {\em preconditions}, {\em negative}, and {\em positive} effects of an action schema $\xi$. Our encoding exploits the syntactic constraint of \strips\ so, if {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a negative effect in $\xi$ while if $pre\_e\_\xi$ does not hold but {\tt\small eff\_e\_$\xi$} holds, it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a positive effect in $\xi$. Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema.
      \end {itemize}

@@

      \item {\em Editable} actions whose semantics is given by the value of {\tt\small pre\_e\_$\xi$}, {\tt\small eff\_e\_$\xi$} fluents at the current state. Figure~\ref{fig:editable} shows the PDDL encoding of an {\em editable} {\tt\small stack(?v1,?v2)} schema. Note that this editable schema when the set of fluents {\tiny\tt (pre\_holding\_v1\_stack) (pre\_clear\_v2\_stack) (eff\_holding\_v1\_stack) (eff\_clear\_v2\_stack) (eff\_clear\_v1\_stack) (eff\_handempty\_stack) (eff\_on\_v1\_v2\_stack)} hold at the current state, then it behaves exactly as the original PDDL schema defined in Figure~\ref{fig:propositional}. Formally, given an operator schema $\xi\in\mathcal{M}$ its {\em editable} version is:
\begin{small}  
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi}})=&\{pre\_e\_\xi\implies e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}\\
\cond(\mathsf{editable_{\xi}})=&\{pre\_e\_\xi, eff\_e\_\xi\}\rhd\{\neg e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}},\\
&\{\neg pre\_e\_\xi, eff\_e\_\xi\}\rhd\{e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}.
\end{align*}
\end{small}

@@

\item The clauses in $\Upsilon$ comprises:
      \begin{enumerate}
      \item The {\em unit clauses} given by the fluents that hold in the initial state $I=s_0$ and $mode_{commit}$ set to true.
      \item The clauses representing that the actual value of fluents {\tt\small pre\_e\_$\xi$}, {\tt\small eff\_e\_$\xi$} is unknown. In other words, that any model from the \strips\ space of models (following the previously mentioned syntactic constraints) can initially be part of the {\em least-commitment} action model. Formally, for every $\xi$ and $e\in{\mathcal I}_{\Psi,\xi}$, then $\Upsilon$ includes these two clauses:
            \begin{itemize}
            \item {\tt\small pre\_e\_$\xi$} $\vee$ {\tt\small $\neg$pre\_e\_$\xi$}.
            \item {\tt\small eff\_e\_$\xi$} $\vee$ {\tt\small $\neg$eff\_e\_$\xi$}.
            \end{itemize}
      \end{enumerate}
One can also add here clauses that encode {\em domain-specific constraints} (as mentioned in the previous section) to make the conformant planning problem easier to be solved for a specific domain. 
\item The new goals are $G_c=s_m\cup\{test_m\}$.
\end{itemize}

@@

\subsection{Optimization of the compilation}
In fact, we do not need to compute the {\em least-commitment} action model (i.e. solve the $P_c$ conformant planning problem) from scratch for all the $\tup{s_0^o,s_1^o \ldots , s_m^o}$ observations. A more efficient approach is to compute the {\em least-commitment} action model from scratch only for the first observation, that is for $\tup{s_0^o,s_1^o}$. Then, the obtained {\em least-commitment} action model can be used as input for improving the computation of the two first observations $\tup{s_0^o,s_1^o,s_2^o}$. In other words, including the commits found in the previous step as new {\em unit clauses} in set of the {\em initial clauses} $\Upsilon$. This process is repeated until the {\em least-commitment} action model is computed  for the full sequence of state observations $\tup{s_0^o,s_1^o \ldots , s_m^o}$.

@@

\section{Conclusions}
Related work~\cite{SternJ17}.

%% The file named.bst is a bibliography style file for BibTeX 0.99c
\bibliographystyle{named}
\bibliography{planlearnbibliography}
